Applications of Time-Inhomogeneous Multi-State Models (including Bayesian statistics and swine flu)

### 2009-10-23 docs/pandemic/bayesPandemic.rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@431 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[xcolor=dvipsnames]{beamer}
\usepackage[nogin]{Sweave}

@@

\title{Bayesian statistics and  swine flu}
\author{Patrick Brown$^1$,  Peter Neal$^2$}
\institute{1:Cancer Care Ontario and Dalla Lana School of Public Health, University of Toronto; 2: School of Mathematics, University of Manchester} 
\usepackage{amsmath,subfigure,alltt}
\setkeys{Gin}{width=\columnwidth}

@@

\end{document}

D75A0BD6A6A1

>>>>

\documentclass[xcolor=dvipsnames]{beamer}

\usepackage[nogin]{Sweave}
\usepackage{tikz,pgflibraryshapes}

@@

\title{Applications of Time-Inhomogeneous Multi-State Models (including Bayesian statistics and swine flu)}
\author{Patrick Brown$^1$,  Peter Neal$^2$}
\institute{1:Cancer Care Ontario and Dalla Lana School of Public Health, University of Toronto; 2: School of Mathematics, University of Manchester} 
\usepackage{amsmath,subfigure,alltt}
\setkeys{Gin}{width=\columnwidth}

@@

\begin{frame}
\frametitle{Multi-state models}
\begin{itemize}
	\item The Swin flu application was \emph{ordinal}
	\item Once an individual makes a transtion they can't go back
	\item but what if they can go back?
\end{itemize}

\begin{frame}
\frametitle{Endemic diseases in Tanzanian cows}
\begin{columns}
\column{0.5\textwidth}
\includegraphics{Figures/G-cowTS}
\column{0.5\textwidth}
\begin{itemize}
	\item Sampled up to 11 different occasions
	\item Approximately 6 weeks apart
	\item 60 farms, 100 animals
\end{itemize}
\end{columns}
\end{frame}

\begin{frame}
\frametitle{Questions}
\begin{itemize}
	\item How long does an episode last?
	\item How does age affect infection probabilities?
	\item Do cows located close together have similar infection probabilities?

\begin{itemize}
	\item Transmission between animals?
 \item A spatially varying risk factor (density of ticks)?
\end{itemize}
\item Note: We don't observe infection and recovery times.
\item We only know infection status on certain days.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{A Susceptible-Infected-Susceptible model}
\begin{tikzpicture}[fill=blue!20,line width=2pt,shorten >=5pt]
\tikzstyle{every node}=[draw,fill,ellipse,minimum height=2cm,minimum width=3cm] 
\path (0,0) node(S) {Susceptible}
	 		(5,0) node(I)  {Infected};
\draw[blue,->](S) .. controls +(2,2) and +(-2,2)  ..(I);
\draw[blue,->](I) .. controls +(-2,-2) and +(2,-2)  ..  (S);
\end{tikzpicture}
\begin{itemize}
	\item Every animal alternates between two states
	\item Infections happen at rate $\delta$
	\item Recoveries have rate $\rho$
\end{itemize}
\end{frame}
\begin{frame}

\frametitle{Infection probability changes with age}
\begin{tikzpicture}[fill=white,font=\sffamily]

\begin{scope}[line width=2pt,->]
\draw(0,0) -- (0,5) node[pos=0.5,sloped,above=7pt]{$\delta$(age)};
\draw(0,0) -- (8,0) node[pos=0.5,sloped,below=7pt]{age};
\end{scope}

\draw (0.5,4) -- (2.5,1) --  (8,1);

\begin{scope}[dotted,line width=0.5pt]
\draw (2.5,0) -- (2.5, 1) node[below,pos=0]{$\tau_0$};
\end{scope}

\end{tikzpicture}
\begin{itemize}
	\item Recovery rate $\rho$ is constant
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Model}
Animal $i$ located at $s_i$ has infection rate
\[
\delta_{i}(t) =\begin{cases} \exp(\delta_0 + X_{i}' \beta + w_i \gamma + A(s_i)) & w_i(t) < \tau_0\\
\exp(\delta_0 + X_{i}' \beta + w_i \tau_0 +A(s_i)) & w_i(t) \geq \tau_0 \end{cases} 
\]
\begin{itemize}
	\item $X_{i}$ is indicator variables denoting if the animal grazes, is sprayed, or eats grass
	\item $w_i(t)$ is the age of animal $i$ at time $t$.
	\item $A(\cdot)$ is a Gaussian random field with covariance function 
	\[
	\text{cov}(A(s_0), A(s_0+u)) = \sigma^2 \exp(-||u||/\phi)
	\]
	\item Recovery rate is $\rho$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Inference with constant rates}
\begin{itemize}
	\item Alternating states model, see Grimmett and Stirzaker.
	\item Transition model.
	\item To construct the likelihood, we need the transition probability matrix:
	\[
	P_u = \left(\begin{array}{ll} Pr(Y_{t+u} = 0 | Y_t = 0) &  Pr(Y_{t+u} = 1 | Y_t = 0)\\
 Pr(Y_{t+u} = 0 | Y_t = 1) &  Pr(Y_{t+u} = 1 | Y_t = 1) \end{array}\right) 
	\]
	\item Cow $i$ has infection rate $\delta_i$ and recovery rate $\rho$ 
	\begin{align*}
	G_i = &\left(\begin{array}{cc} -\delta_i & \delta_i \\ \rho & -\rho  \end{array}\right) \Leftarrow \text{Generating Matrix}
\\
P_u = &\exp(u G_i)
\end{align*}
\item This is a matrix exponential $\exp(X) = \sum_{n=1}^\infty X^n/n!$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Probabilities}
\begin{itemize}
	\item Unconditional distribution
	\[
	 Pr(Y_{it} = 1) = \frac{\delta_i}{\delta_i + \rho} 
	\]
	\item Susceptible $\rightarrow$ infected
	\[
	  Pr(Y_{i,t+u} = 1 | Y_{it} = 0) = \frac{\delta_i}{\delta_i + \rho} \left( 1- \exp[-u(\delta_i + \rho)]\right)
	\]
	\item Infected $\rightarrow$ infected
	\[
	  Pr(Y_{i,t+u} = 1 | Y_{it} = 1) = \frac{\delta_i}{\delta_i + \rho} + 
	  	\frac{\rho}{\delta_{i} + \rho} \exp[-u(\delta_{i} + \rho)]
	  \]
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Notes}
\begin{itemize}
	\item The observations are Bernoulli random variables, with a peculiar formula for calculating the probabilities
	\[
	p_{it} = pr(Y_{it} = 1) = f(X_i, Y_{i,t-1})
	\]
	\item The likelihood function is a product of the probabilities (or 1 minus them).
	\[
	L(\delta_0, \rho, \beta ) =  \prod_{it : Y_{it}=1}p_{it} \prod_{it : Y_{it}=0}(1-p_{it} )
	\]
	\item  The Markov assumption is why this problem is tractable
	\item Markov property: the future depends only on the present, not the past.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{More  Notes}
\begin{itemize}
\item Number of infections in an interval of length $u$ $\sim \text{Poisson}(u \delta)$
\item Duration of a disease episode $\sim \text{Exponential}(1/\rho)$
\item The Exponential-Poisson combination is a result of assuming the Markov property.
\end{itemize}
\end{frame}

\frametitle{Changing rates?}
\begin{itemize}
\item A surprisingly difficult problem
	\item Peter Neal (University of Manchester) found the solution:
\end{itemize}
\begin{align*}
G_i(t)= &\begin{pmatrix} -\exp(\delta_0 + \beta t) & \exp(\delta_0 + \beta t) \\ \rho &-\rho  \end{pmatrix}\\
P(t, t+u) = & \exp\left( \int_0^{u}  G(t+x) dx\right)\\
\end{align*}
\pause
Later, Peter says: \emph{The above expression is unfortunately wrong.  The problem being that matrix multiplication is non-commutative.  In other words $e^Ae^B \neq e^{A+B}$ unless $AB=BA$.  }

\begin{frame}
\frametitle{Solutions}
\begin{block}{Brute force}
\begin{itemize}
	\item Reversible jump MCMC, data augmentation
	\item number of transitions is a latent variable, augmented
	\item locations of those transitions is also augmented
\end{itemize}
\end{block}
\begin{block}{Approximations}
\begin{itemize}
	\item Peter's found upper and lower bounds to the probabilities
	\item When updating, check if proposal is above upper or below lower bound
	\item If not, recalculate boundaries with greater precision
\end{itemize}

\begin{frame}
\frametitle{Local Likelihood}
with Gerald Lebovic

\begin{itemize}
\item Write $Y_{ij}$ as the state of individual $i$ at time $T_{ij}$.
	\item To estimate $\delta(t)$, use $\{Y_{ij} |t-h< T_{ij} < t+h \}$
	\item or some weighted kernel kernel
	\item suppose data are unbalanced?
	\item There may be no observations in the interval
	\item but observations just outside could be informative
	\item with a kernel, observations might be in places of low weight, but be informative about the (missing) data with high weight.
\end{itemize}

\begin{frame}
\frametitle{Local EM}
\begin{itemize}
	\item Put the data on a fine grid, $\tilde Y_{ik}, \tilde T_{ik}$, $k=1 \ldots K$
	\item Most of the $\tilde Y_{ik}$ are missing
\item Start with an estimate $\hat\delta^{(0)}(t)$:
\end{itemize}
\begin{description}
\item[E step]: compute $\text{E}[\tilde Y_{ik}|Y;\hat\delta^{(p)}(t)]$
\item [M step]: use local likelihood to estimate $\hat\delta^{(p+1)}(t)$
\end{description}

\begin{frame}
\frametitle{What is $\tilde Y_{ik}$?}
\begin{itemize}
	\item If three states, $\tilde Y_{ik} = (p_{ik1}, p_{ik2}, p_{ik1})$, probability of being in each of the three states
	\item No\ldots, not easy to calculate the likelihood from these data
	\item $\tilde Y_{ik}$ is a 3x3 matrix:
\[
\tilde Y_{ik}(a,b) = pr(Y_{ik-1} = a, Y_{ik}=b)
\] 
\item If $Y_{ik}$ and $Y_{ik-1}$ are known, one ``1'' and zeros
\item If one of $Y_{ik}$ and $Y_{ik-1}$ are known, one non-zero row or column
\item then if $P_k$ is the transition matrix 
\[
\log[L(\tilde Y)] = \sum_{i,k} \tilde Y_{ik}' P \tilde Y_{ik}
\]
\end{itemize}

\begin{frame}
\frametitle{Another example: smoking}
\includegraphics[width=2in]{Figures/G-ukDataCurrent.pdf}\includegraphics[width=2in]{Figures/G-ukDataFormer.pdf}

Current smokers \hspace{1.5in} Never smoked
\end{frame}

\begin{frame}
\frametitle{A simple model}
\begin{tikzpicture}[fill=white,line width=2pt,shorten >=5pt,font=\sffamily\small]

\begin{scope}
\tikzstyle{every node}=[draw,fill,ellipse,minimum height=1.4cm,minimum width=1.5cm]

% the nodes
\node (N) at (0,0) {Never Smoked};
\node (C)	at (4,0)  {Current Smoker};
\node	(F) at (8,0)  {Former Smoker};

\end{scope}

% paths between the nodes
\draw[black,->] (N) .. controls +(1.5,1.5) and +(-1.5,1.5)  .. node[below]{start} (C);
\draw[black,->](C) .. controls +(1.5,1.5) and +(-1.5,1.5)  .. node[below]{quit}(F);

\end{tikzpicture}
\end{frame}

\begin{frame}
\frametitle{A more complicated model}

\hspace*{-0.6in}\begin{tikzpicture}[fill=white,line width=2pt,shorten >=5pt,font=\sffamily]

\begin{scope}
\tikzstyle{every node}=[draw,fill,ellipse,minimum height=1.4cm,minimum width=1.5cm]

% the nodes
\node (N) at (0,0) {Never Smoked};
\node (C)	at (5,0)  {Current Smoker};
\node	(F) at (10,0)  {Former Smoker};
\node	(D) at (5,3)  {Death};

\end{scope}

% paths between the nodes
\draw[black,->] (N) .. controls +(1.5,1.5) and +(-1.5,1.5)  .. node[below]{start} (C);
\draw[black,->](C) .. controls +(1.5,1.5) and +(-1.5,1.5)  .. node[below]{quit}(F);
\draw[black,->](F) .. controls +(-1.5,-1.5) and +(1.5,-1.5)  .. node[above]{lapse}(C);
\draw[black,->](F) .. controls +(-0.5,-2.5) and +(0.5,-2.5)  .. node[above]{reclassify}(N);
\draw[black,->](N)..controls + (0, 1.5) .. (D);
\draw[black,->](C)--(D);
\draw[black,->](F).. controls + (0, 1.5) ..(D);

\end{tikzpicture}

\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Transition models can model complex correlation structures with a small number of parameters
\item Extensions to time-varying rates are useful but non-trivial
\item Since these models make scientific sense, they're easier to build and interpret
\item Possible extensions
\begin{itemize}
\item Weibull distribution for disease episode length?
\item Non-linear rate functions?
\item Likely to be impossibly complicated.
\end{itemize}
\pause
\item Don't dismiss stochastic processes as unnecessarily complex and of little use!
\end{itemize}

\end{frame}
\end{document}
